Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2019-04-13 article.tex
Udate the manuscript to the newest dMod version

Although the problem of parameter estimation in non-linear dynamic systems is highly relevant and at the heart of statistical computing, to this day there are not more than four \proglang{R} packages published on the topic on the Comprehensive \proglang{R} Archive Network (CRAN), namely \pkg{FME} \citep{FME}, \pkg{nlmeODE} \citep{nlmeODE}, \pkg{mkin} \citep{mkin} and \pkg{scaRabee} \citep{scaRabee}. All packages have in common that they are built upon the \pkg{deSolve} package \citep{deSolve}.
The packages \pkg{mkin} and \pkg{FME} support ODEs defined by compiled code where \pkg{mkin} also provides tools to autogenerate the \proglang{C} code and compile it using the \pkg{inline} package. For model fitting and uncertainty analysis, \pkg{mkin} fully resorts to the functionality of \pkg{FME}.
Concerning model fitting, \pkg{FME} and \pkg{nlmeODE} (with \pkg{nlme} in the background) use deterministic derivative-based optimizers by default, i.e.~either Levenberg-Marquardt or Newton methods provided by \code{nls.lm()}, \code{optim()} or \code{nlmimb()}. Although all these optimizers support gradient or Hessian information as input, only the \pkg{nlmeODE} package provides an option to augment the ODE by its sensitivity equations to generate derivates for the residuals. By default, sensitivities are computed by finite differences. Last but not least, the \pkg{scaRabee} package uses the Nelder-Mead optimization algorithm which is derivative-free but needs in general more iterations until convergence compared to derivative-based methods.
In all packages, uncertainty analysis is by default based on the variance-covariance matrix, i.e.~on the inverse Hessian matrix of the least squares function. For non-linear models, this method provides a good approximation only if parameters are identifiable and the data is highly informative. If these conditions are not met, more sophisticated methods like e.g.~Markov chain Monte Carlo (MCMC) sampling, implemented in \pkg{FME}, are required.
The strength of \pkg{scaRabee} and especially \pkg{nlmeODE} is multi-conditional fitting. This means that the same model with the same parameters but different forcings to reflect experimental conditions is fitted simultaneously to the condition-specific data sets. In the context of mixed-effects modeling as provided by \pkg{nlmeODE}, parameters can be grouped in fixed effects (same parameters over all conditions) and random effects (parameters are different between conditions).

In this paper we present \pkg{dMod}, an \proglang{R} package on dynamic modeling and parameter estimation. 
The package is grown over the last years and reflects many developments and lessons learned from our research projects with time-resolved experimental data from Systems Biology. They form the basis of the development of \pkg{dMod} and are mostly referenced in the following paragraphs.\\
The aim and core functionality of \pkg{dMod} is
\begin{enumerate}[(i)]
	\item facilitated set-up of dynamic models with automated \proglang{C} code generation for fast simulation of model predictions and model sensitivities,
	\item flexible set-up of general parameter transformations (explicit or implicit) and observation functions, allowing for the implementation of multiple experimental conditions similar to mixed-effects modeling,
	\item parameter estimation based on trust-region optimization of the negative log-likelihood, making use of the sensitivity equations of the dynamic system and of symbolic derivatives of the observation- and parameter transformation functions,
	\item identifiability and uncertainty analysis based on the profile-likelihood method to determine confidence intervals for parameters and predictions.
\end{enumerate}

The packages \pkg{mkin} and \pkg{FME} support ODEs defined by compiled code while \pkg{mkin} also provides tools to autogenerate the \proglang{C} code and compile it using the \pkg{inline} package \citep{inline}. For model fitting and uncertainty analysis, \pkg{mkin} fully resorts to the functionality of \pkg{FME}.
Concerning model fitting, \pkg{FME} and \pkg{nlmeODE} (with \pkg{nlme} in the background) use deterministic derivative-based optimizers by default, i.e.~either Levenberg-Marquardt or Newton methods provided by \code{nls.lm()}, \code{optim()} or \code{nlmimb()}. Although all these optimizers support gradient or Hessian information as input, only the \pkg{nlmeODE} package provides an option to augment the ODE by its sensitivity equations to generate derivates for the residuals. By default, sensitivities are computed by finite differences. Last but not least, the \pkg{scaRabee} package uses the Nelder-Mead optimization algorithm which is derivative-free but generally needs more iterations until convergence compared to derivative-based methods.
In all packages, uncertainty analysis is by default based on the variance-covariance matrix, i.e.~on the inverse Hessian matrix of the least squares function. For non-linear models, this method provides a good approximation only if parameters are identifiable and the data is highly informative. If these conditions are not met, more sophisticated methods like e.g.~Markov chain Monte Carlo (MCMC) sampling, implemented in \pkg{FME}, are required.
The strength of \pkg{scaRabee} and especially \pkg{nlmeODE} is multi-conditional fitting. This means that the same model with the same parameters but different forcings to reflect experimental conditions is fitted simultaneously to the condition-specific data sets. In the context of mixed-effects modeling as provided by \pkg{nlmeODE}, parameters can be grouped in fixed effects (parameters are the same in all conditions) and random effects (parameters are different between conditions).

@@

	\begin{scriptsize}
		$^1$Only ODE, no sensitivity equations generated. $^2$Mixed-effects modeling allows to define parameters in a condition-specific manner. $^3$Other methods used, e.g.~the collinearity method \citep{brun2001practical}. 
	\end{scriptsize}
\end{table}

@@

The implementation of capabilities to generate and propagate derivatives on a compiled-code level distinguishes \pkg{dMod} from other modeling frameworks as discussed above. Most of the standard optimization routines implemented in \proglang{R} need derivative information. However, the computation of derivatives in the context of ODE models holds some pitfalls. The accuracy of sensitivities obtained from finite differences can be insufficient because the step control of the integrator presents an additional source of numeric inaccuracy. Even for numeric methods circumventing this problem, e.g.~complex-step derivatives \citep{squire1998using}, the use of sensitivity equations is still beneficial judging from the accuracy vs.~computational cost ratio. See \cite{raue2013lessons} for a comparison of methods.

@@

\subsection{Dynamic models and model sensitivities}
Dynamic models describe systems with states $x$, usually quantifying the involved species, their interaction and evolution over time. %  as e.g. given by chemical reaction networks.
The time evolution of the states is expressed via a set of ODEs, $\dot x = f(x)$. Although constituting quite a special class of dynamic systems, chemical reaction networks formulated by the law of mass action as considered here allow for surprisingly general applications. Typical ODE examples derived from the law of mass action are:
\begin{itemize}
	\item $\dot x_A = k_p$, constant production of $x_A$ ($\emptyset \rightarrow A$)
	\item $\dot x_C = k_c \,x_A \,x_B = - \dot x_A = - \dot x_B$, complex formation ($A+B \rightarrow C$)
	\item $\dot x_C = - k_d \,x_C$, proportional degradation of $x_C$ ($C \rightarrow \emptyset$)\,.
\end{itemize}
A general chemical reaction 
\begin{align}
	\nu_1 X_1 + \dots + \nu_n X_n \longrightarrow \nu'_1 X_1 + \dots + \nu'_n X_n
	\label{eq:ma_chemical}
\end{align}
for species $X_j$, $j = 1, \dots, n$, and stoichiometric coefficients $\nu_j, \nu'_j\in\mathbb N$ translates into the ODE
\begin{align}
	\dot x_j = \underbrace{k_0 \left(\prod x_i^{\nu_i}\right)}_{w_0(x, k_0)} (\nu_j' - \nu_j),
	\label{eq:ma}
\end{align}
where $k_0$ is the reaction rate.  When several reactions $1, \dots, r$ are involved, all contributions have to be summed up which can be expressed by the matrix equation
\begin{align}
	\dot x = S w(x, k).
	\label{eq:Sv}
\end{align}
%$\dot x_l = \sum_{j = 1\dots m}  k_{j} \prod_{i = 1\dots n} \nu_{i,j} x_i^{\nu_{i,j}}$ for a system of $n$ states and $m$ reactions with rate constants $p = \{ k_j, j = 1\dots m\}$ and stoichiometric constants $\nu_{i,j} \in \mathbb{Z}$.
%An even more general expression of the system is given by $\dot x = S v(x,p)$
Here, $S$ denotes the $n\times r$ stoichiometry matrix where each column corresponds to one reaction and the coefficients within one column are the values $\nu'_j - \nu_j$ of the corresponding reaction. The vector $w(x, k)$ denotes the vector of $r$ reaction fluxes where the vectors $x$ and $k$ correspond to states and rate constants, respectively.
%containing the stoichiometric constants $\nu_{i,j}$
%obtained from the collection of vectors $\nu' - \nu$ for reactions $1, \dots, r$,
%and the flux vector $w(x,k)\in\mathbb R^{r}$.
Eq.~\eqref{eq:Sv} also holds for general ODEs that do not follow from mass action kinetics. Moreover the system may be explicitly time-dependent and may contain forcings $u(t)$ which e.g.~describe the external stimulation of the system. A general form of the dynamic model is therefore given by
\begin{align}
	\dot x = f(x,k,u,t) = S w(x,k,u,t), \quad\quad \textrm{with }  x(t = 0) = x_0.
\end{align}
The solution $x(t,\theta)$ for a given parameter vector $\theta = (k, x_0)$ is called model prediction. %The parameters $\theta$ contain the rate constants, initial values $x_0$ and additional parameters included in the forcings.
It is generally assumed that the stoichiometry matrix $S$ of the system is known.
Besides the model predition itself, also the sensitivity of the prediction to changes in the parameter values is of interest.
The sensitivities $s_i = \frac{\partial x}{\partial \theta_i}$ satisfy the \textit{sensitivity equations}
\begin{align}
	\dot s_i  = \frac{\partial f}{\partial x} s_i + \frac{\partial f}{\partial \theta_i}, \qquad \textrm{with }s_i(t = 0) = \frac{\partial x_0}{\partial \theta_i},
\end{align}
a system of ordinary differential equations that in general depend on the states $x$ and, therefore, need to be solved jointly with the original ODE.

@@

Parameter estimation is a common task in statistics. It describes the process of inferring parameter values or parameter ranges of a statistical model based on observed data. Over decades, appropriate estimators have been developed for different problem classes. The principle of maximum likelihood allows to derive an estimator which is especially suited for applications where the distribution of the measurement noise is known. This knowledge about the structure of the noise constitutes additional information that makes the maximum-likelihood estimator (MLE) \textit{efficient}, i.e.~from all unbiased estimators the MLE has the lowest variance. Other properties of maximum-likelihood estimation are \textit{consistency}, i.e.~the estimated parameter value approaches the true parameter value in the limit of infinite data-sample size, and \textit{asymptotic normality}, i.e.~in the limit of infinite data the MLE follows a multivariate normal distribution. See \cite{azzalini1996statistical} for an introduction to likelihood theory.

@@

Optimization by the Newton method attempts to iteratively find the root of the gradient $\nabla l (\theta)$ by the recursion
\begin{align}
	\theta^{(n+1)} = \theta^{(n)} - Hl(\theta)^{-1}\nabla l(\theta),
	\label{eq:newton}
\end{align}
where $Hl(\theta)$ denotes the Hessian, i.e.~the matrix of second derivatives of $l(\theta)$. The fact that for normally distributed noise the log-likelihood is a least squares function, has a big advantage: gradient and (approximate) Hessian can be computed from first-order derivatives  \citep{press1996numerical}. Since the second-order contributions to the Hessian can be neglected, gradient and Hessian read
\begin{align}
	\nabla l(\theta) = 2 r(\theta) J(\theta) \quad\quad Hl(\theta) \approx 2 J(\theta)^T J(\theta),
	\label{eq:derivatives}
\end{align}
with the weighted residual vector $r_i(\theta) = \frac{y_i(\theta) - y_i^D}{\sigma_i}$ and its first derivative, the Jacobian matrix
\begin{align}
	J_{ij}(\theta) = \frac{\partial r_i}{\partial \theta_j}(\theta) = \frac{1}{\sigma_i} \frac{\partial y_i}{\partial \theta_j}(\theta). \label{eq:jacobian}
\end{align}
As indicated by eqs.~\eqref{eq:derivatives}-\eqref{eq:jacobian},  first order model sensitivities $\frac{\partial y_i}{\partial\theta_j}$ are sufficient to compute gradient and Hessian in good approximation.

The Newton recursion, eq.~\eqref{eq:newton}, converges in one iteration if $l(\theta)$ is a quadratic form. However, $l(\theta)$ is only quadratic if the model $y(\theta)$ is linear in $\theta$ which is usually not the case for ODE models. On the other hand, each smooth function, no matter if quadratic or not, can be approximated by a quadratic function based on its Taylor series.
%Depending on the function's shape, the \textit{trust region}, i.e.~the region where $l(\theta)$ is well approximated, can become quite small. Accordingly, a Newton step pointing far beyond this region could be misleading and not helpful for optimization. This is the idea behind trust-region optimization \citep{wright1999numerical}.
Thus, $l(\theta)$ can be approximated by a quadratic function in at least a small region around $\theta$. The idea of trust-region optimization is to confine a Newton step to a region, the \textit{trust region}, where the quadratic approximation holds \citep{wright1999numerical}.
In each iteration, the trust-region radius is adjusted and the optimization problem restricted to the trust region is solved. For parameter estimation in ODE models this has the additional advantage that parameter changes are constricted and optimization steps into parameter regions where the ODE cannot be numerically solved any more  become less frequent. This makes trust-region optimization the method of choice for \pkg{dMod}.

@@

\pkg{dMod} provides several ways to define the differential equations. An \code{eqnvec} of equations can be explicitly formulated, analogously to a \code{c()} command. Especially for chemical reactions it can be tedious to keep track of all gain and loss terms. Therefore, \pkg{dMod} provides also the \code{eqnlist} class which encodes the ODE by a list of (1) state names, (2) rate expressions, (3) compartment volumes and (4) the stoichiometric matrix. Each reaction flux, c.f.~eq.~\eqref{eq:Sv}, occurs only once and the gain and loss is represented by the coefficients in the stoichiometric matrix. \pkg{dMod} supports the read-in of a csv file with the stoichiometric matrix and rate expressions and provides a function \code{addReaction()} to construct an \code{eqnlist} object step-by-step or add further reactions to an already existing \code{eqnlist}. An \code{eqnlist} is converted to (differential) equations, i.e.~an \code{eqnvec}, by \code{as.eqnvec()}. The conversion does all the bookkeeping of gain and loss terms and volume ratios due to compartment transitions.

@@

\pkg{dMod} seeks to stay close to the mathematical formulation, i.e.~\code{x <- Xs(myodemodel)} will indeed return an \proglang{R} function, an object of class \code{prdfn}, which expects arguments \code{times} and \code{pars} and turns them into a model prediction. The letter \code{"s"} in \code{"Xs"} refers to sensitivities, i.e.~the solution $x(t)$ is returned, the sensitivities $\frac{\partial x}{\partial p}(t)$ are returned, too.

Besides parameters, the prediction might also depend on forcings $u(t)$ and sudden events, e.g.~setting states to 0 at a predefined point in time. Since forcings and events are by default fixed, they are defined together with the  prediction function, \code{x <- Xs(odemodel, forcings = myforcings, events = myevents)}. The definitions of both, forcings and events, are compatible with the way they are defined in \pkg{deSolve}.

@@

It is one of the fundamental concepts of \pkg{dMod} to allow concatenation of functions via the \code{"*"} operator. The mathematical formulation $y = g\circ x$, i.e.~$y(t, p) = (g\circ x)(t, p):= g(x(t, p), p, t)$ becomes \code{y = g*x} in \proglang{R}. To obtain derivatives $\frac{\partial y}{\partial p}$, the chain rule is applied: $\frac{\partial y}{\partial p} = \frac{\partial g}{\partial x}\frac{\partial x}{\partial p} + \frac{\partial g}{\partial p}$. This means that \code{Y()} needs to be informed which symbols are states and parameters to generate the corresponding expressions $\frac{\partial g}{\partial x}$ and $\frac{\partial g}{\partial p}$. The observation function \code{g} takes care of computing theses derivatives from the symbolic expressions and doing the matrix multiplication with the sensitivities $\frac{\partial x}{\partial p}$ from the prediction function. Evaluation of symbolic expressions can become inefficient in \proglang{R}. Therefore, the observation function is usually translated into a \proglang{C} code and compiled.

When observation- and prediction functions are concatenated, the result is a prediction function, e.g.~\code{y = g*x} is the \proglang{R} function computing values $y(t)$ from the arguments \code{times} and \code{pars} via evaluation of the ODE and subsequent evaluation of the observation function. Two observation functions can be concatenated, too, again yielding an observation function.

@@

Similar to prediction- and observation functions, parameter functions not only return parameter values but the Jacobian of the transformation, too. Exploiting the chain rule, the derivatives are propagated, allowing to define \code{y = g*x*p} which is a function returning $y(t, \theta)$ and $\frac{\partial y}{\partial \theta}(t)$, i.e.~\code{g*x*p} is a prediction function.

@@

The aim of \pkg{dMod} is to allow for ``simultaneous'' predictions under several conditions and compare these predictions to the corresponding experimental data sets to estimate model parameters.
Different experimental conditions are typically expressed by the fact that some parameters are different between conditions whereas others are common to all conditions. This situation occurs e.g.~if perturbation experiments are performed, affecting only few parts of a system.
%The full information contained in several perturbation experiments is only used if the multi-conditional model prediction is set up in a way that all model parameters not being changed from condition to condition correspond to exactly one parameter. On the other hand, parameters which are systematically perturbed correspond to different parameters.
Mathematically speaking, we want to construct a parameter transformation
\begin{equation}
	\begin{aligned}
		\Phi:\quad \mathbb R^p & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^q \\
		\theta & \longmapsto  \left(\Phi_1(\theta), \dots, \Phi_n(\theta)\right)
	\end{aligned}
	\label{eq:partrafo}
\end{equation}
where $i = 1, \dots, n$ corresponds to the different conditions. If all parameters are shared throughout all conditions, then $p = q$. If, however, all parameters are distinct, then $p = n\cdot q$. Perturbation experiments correspond to a situation where $p\gtrsim q$.

@@

Prediction- and observation functions are generalized to multiple conditions by the \code{"+"} operator, too. Mathematically speaking, they become
\begin{align*}
	\begin{aligned}
		x: &  & \mathbb R \times \bigoplus_{i = 1}^n \mathbb R^q & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^m \\
		 & & (t, p_1, \dots, p_n) & \longmapsto (x_1(t, p_1), \dots, x_n(t, p_n)) \\
		g: &  & \bigoplus_{i = 1}^n \mathbb R^m \times \bigoplus_{i = 1}^n \mathbb R^q\times \mathbb R& \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^s \\
		& & (x_1(t), \dots, x_n(t), p_1, \dots, p_n, t) & \longmapsto (g_1(x_1(t), p_1, t), \dots, g_n(x_n(t, p_n), p_n, t)).
	\end{aligned}
	\label{}
\end{align*}
In words, if prediction- or observation functions are defined for different conditions then they expect condition-specific inputs which are evaluated by the matching functions. Examples for condition-specific prediction functions typically involve different forcings or events. Observation functions can e.g.~differ between different measurement techniques. In all these cases, the different prediction functions $x_1, \dots, x_n$ or observation functions $g_1, \dots, g_n$ are defined, referencing the condition, and combined by the \code{"+"} operator analogously to \code{p}.
% Aufruf der Funktionen mit subset von conditions

@@

# Translate into ODE model
mymodel <- odemodel(reactions, modelname = "bamodel")
\end{CodeInput}
\end{CodeChunk}

>>>>

Although the problem of parameter estimation in non-linear ODE models is highly relevant and at the heart of statistical computing, there are currently not more than four \proglang{R} packages published on the topic on the Comprehensive \proglang{R} Archive Network (CRAN), namely \pkg{FME} \citep{FME}, \pkg{nlmeODE} \citep{nlmeODE}, \pkg{mkin} \citep{mkin} and \pkg{scaRabee} \citep{scaRabee}. All packages have in common that they are built upon the \pkg{deSolve} package \citep{deSolve}. Taking a broader perspective on the topic of dynamic modeling and inference, we find more packages, dealing with, e.g., discrete-time and continuous-time stochastic systems \citep{CollocInfer} or the statistical inference of partially observed Markov processes \citep{pomp}.

The packages \pkg{mkin} and \pkg{FME} support ODEs defined by compiled code while \pkg{mkin} also provides tools to autogenerate the \proglang{C} code and compile it using the \pkg{inline} package \citep{inline}. For model fitting and uncertainty analysis, \pkg{mkin} fully resorts to the functionality of \pkg{FME}.
Concerning model fitting, \pkg{FME} and \pkg{nlmeODE} (with \pkg{nlme} in the background) use deterministic derivative-based optimizers by default, i.e., either Levenberg-Marquardt or Newton methods provided by \code{nls.lm()}, \code{optim()} or \code{nlmimb()}. Although all these optimizers support gradient or Hessian information as input, only the \pkg{nlmeODE} package provides an option to augment the ODE by its sensitivity equations to generate derivates for the residuals. By default, sensitivities are computed by finite differences. Last but not least, the \pkg{scaRabee} package uses the Nelder-Mead optimization algorithm which is derivative-free but generally needs more iterations until convergence compared to derivative-based methods.
In all packages, uncertainty analysis is by default based on the variance-covariance matrix, i.e., non the inverse Hessian matrix of the least squares function. For non-linear models, this method provides a good approximation only if parameters are identifiable and the data is highly informative. If these conditions are not met, more sophisticated methods like, e.g., Markov chain Monte Carlo (MCMC) sampling, implemented in \pkg{FME}, are required.
The strength of \pkg{scaRabee} and especially \pkg{nlmeODE} is multi-conditional fitting. This means that the same model with the same parameters but different forcings to reflect experimental conditions is fitted simultaneously to the condition-specific data sets. In the context of mixed-effects modeling as provided by \pkg{nlmeODE}, parameters can be grouped in fixed effects (parameters are the same in all conditions) and random effects (parameters are different between conditions).

@@

	\begin{scriptsize}
		$^1$Only ODE, no sensitivity equations generated. $^2$Mixed-effects modeling allows to define parameters in a condition-specific manner. $^3$Other methods used, e.g., the collinearity method \citep{brun2001practical}. 
	\end{scriptsize}
\end{table}

@@

The implementation of capabilities to generate and propagate derivatives on a compiled-code level distinguishes \pkg{dMod} from other modeling frameworks as discussed above. Most of the standard optimization routines implemented in \proglang{R} need derivative information. However, the computation of derivatives in the context of ODE models holds some pitfalls. The accuracy of sensitivities obtained from finite differences can be insufficient because the step control of the integrator presents an additional source of numeric inaccuracy. Even for numeric methods circumventing this problem, e.g., complex-step derivatives \citep{squire1998using}, the use of sensitivity equations is still beneficial judging from the accuracy vs.~computational cost ratio. See \cite{raue2013lessons} for a comparison of methods.

@@

\subsection{Dynamic models and model sensitivities}
Dynamic models describe systems with states $x$, usually quantifying the involved species, their interaction and evolution over time. %  as e.g. given by chemical reaction networks.
The time evolution of the states is expressed via a set of ODEs, $\dot x = f(x)$. Although constituting quite a special class of dynamic systems, chemical reaction networks formulated by the law of mass action as considered here allow for surprisingly general applications. Typical ODE examples derived from the law of mass action are:
\begin{itemize}
	\item $\dot x_A = k_p$, constant production of $x_A$ ($\emptyset \rightarrow A$)
	\item $\dot x_C = k_c \,x_A \,x_B = - \dot x_A = - \dot x_B$, complex formation ($A+B \rightarrow C$)
	\item $\dot x_C = - k_d \,x_C$, proportional degradation of $x_C$ ($C \rightarrow \emptyset$)\,.
\end{itemize}
A general chemical reaction 
\begin{align}
	\nu_1 X_1 + \dots + \nu_n X_n \longrightarrow \nu'_1 X_1 + \dots + \nu'_n X_n
	\label{eq:ma_chemical}
\end{align}
for species $X_j$, $j = 1, \dots, n$, and stoichiometric coefficients $\nu_j, \nu'_j\in\mathbb N$ translates into the ODE
\begin{align}
	\dot x_j = \underbrace{k_0 \left(\prod x_i^{\nu_i}\right)}_{w_0(x, k_0)} (\nu_j' - \nu_j),
	\label{eq:ma}
\end{align}
where $k_0$ is the reaction rate.  When several reactions $1, \dots, r$ are involved, all contributions have to be summed up which can be expressed by the matrix equation
\begin{align}
	\dot x = S w(x, k).
	\label{eq:Sv}
\end{align}
%$\dot x_l = \sum_{j = 1\dots m}  k_{j} \prod_{i = 1\dots n} \nu_{i,j} x_i^{\nu_{i,j}}$ for a system of $n$ states and $m$ reactions with rate constants $p = \{ k_j, j = 1\dots m\}$ and stoichiometric constants $\nu_{i,j} \in \mathbb{Z}$.
%An even more general expression of the system is given by $\dot x = S v(x,p)$
Here, $S$ denotes the $n\times r$ stoichiometry matrix where each column corresponds to one reaction and the coefficients within one column are the values $\nu'_j - \nu_j$ of the corresponding reaction. The vector $w(x, k)$ denotes the vector of $r$ reaction fluxes where the vectors $x$ and $k$ correspond to states and rate constants, respectively.
%containing the stoichiometric constants $\nu_{i,j}$
%obtained from the collection of vectors $\nu' - \nu$ for reactions $1, \dots, r$,
%and the flux vector $w(x,k)\in\mathbb R^{r}$.
Eq.~\eqref{eq:Sv} also holds for general ODEs that do not follow from mass action kinetics. Moreover the system may be explicitly time-dependent and may contain forcings $u(t)$ which, e.g., describe the external stimulation of the system. A general form of the dynamic model is therefore given by
\begin{align}
	\dot x = f(x,k,u,t) = S w(x,k,u,t), \quad\quad \textrm{with }  x(t = 0) = x_0.
\end{align}
The solution $x(t,\theta)$ for a given parameter vector $\theta = (k, x_0)$ is called model prediction. %The parameters $\theta$ contain the rate constants, initial values $x_0$ and additional parameters included in the forcings.
It is generally assumed that the stoichiometry matrix $S$ of the system is known.
Besides the model predition itself, also the sensitivity of the prediction to changes in the parameter values is of interest.
The sensitivities $s_i = \frac{\partial x}{\partial \theta_i}$ satisfy the \textit{sensitivity equations}
\begin{align}
	\dot s_i  = \frac{\partial f}{\partial x} s_i + \frac{\partial f}{\partial \theta_i}, \qquad \textrm{with }s_i(t = 0) = \frac{\partial x_0}{\partial \theta_i},
\end{align}
a system of ordinary differential equations that in general depend on the states $x$ and, therefore, need to be solved jointly with the original ODE.

@@

Parameter estimation is a common task in statistics. It describes the process of inferring parameter values or parameter ranges of a statistical model based on observed data. Over decades, appropriate estimators have been developed for different problem classes. The principle of maximum likelihood allows to derive an estimator which is especially suited for applications where the distribution of the measurement noise is known. This knowledge about the structure of the noise constitutes additional information that makes the maximum-likelihood estimator (MLE) \textit{efficient}, i.e., from all unbiased estimators the MLE has the lowest variance. Other properties of maximum-likelihood estimation are \textit{consistency}, i.e., the estimated parameter value approaches the true parameter value in the limit of infinite data-sample size, and \textit{asymptotic normality}, i.e., in the limit of infinite data the MLE follows a multivariate normal distribution. See \cite{azzalini1996statistical} for an introduction to likelihood theory.

@@

Optimization by the Newton method attempts to iteratively find the root of the gradient $\nabla l (\theta)$ by the recursion
\begin{align}
	\theta^{(n+1)} = \theta^{(n)} - Hl(\theta)^{-1}\nabla l(\theta),
	\label{eq:newton}
\end{align}
where $Hl(\theta)$ denotes the Hessian, i.e., the matrix of second derivatives of $l(\theta)$. The fact that for normally distributed noise the log-likelihood is a least squares function, has a big advantage: gradient and (approximate) Hessian can be computed from first-order derivatives  \citep{press1996numerical}. Since the second-order contributions to the Hessian can be neglected, gradient and Hessian read
\begin{align}
	\nabla l(\theta) = 2 r(\theta) J(\theta) \quad\quad Hl(\theta) \approx 2 J(\theta)^T J(\theta),
	\label{eq:derivatives}
\end{align}
with the weighted residual vector $r_i(\theta) = \frac{y_i(\theta) - y_i^D}{\sigma_i}$ and its first derivative, the Jacobian matrix
\begin{align}
	J_{ij}(\theta) = \frac{\partial r_i}{\partial \theta_j}(\theta) = \frac{1}{\sigma_i} \frac{\partial y_i}{\partial \theta_j}(\theta). \label{eq:jacobian}
\end{align}
As indicated by eqs.~\eqref{eq:derivatives}-\eqref{eq:jacobian},  first order model sensitivities $\frac{\partial y_i}{\partial\theta_j}$ are sufficient to compute gradient and Hessian in good approximation.

The Newton recursion, eq.~\eqref{eq:newton}, converges in one iteration if $l(\theta)$ is a quadratic form. However, $l(\theta)$ is only quadratic if the model $y(\theta)$ is linear in $\theta$ which is usually not the case for ODE models. On the other hand, each smooth function, no matter if quadratic or not, can be approximated by a quadratic function based on its Taylor series.
Thus, $l(\theta)$ can be approximated by a quadratic function in at least a small region around $\theta$. The idea of trust-region optimization is to confine a Newton step to a region, the \textit{trust region}, where the quadratic approximation holds \citep{wright1999numerical}.
In each iteration, the trust-region radius is adjusted and the optimization problem restricted to the trust region is solved. For parameter estimation in ODE models this has the additional advantage that parameter changes are constricted and optimization steps into parameter regions where the ODE cannot be numerically solved any more  become less frequent. This makes trust-region optimization the method of choice for \pkg{dMod}.

@@

\pkg{dMod} provides several ways to define the differential equations. An \code{eqnvec} of equations can be explicitly formulated, analogously to a \code{c()} command. Especially for chemical reactions it can be tedious to keep track of all gain and loss terms. Therefore, \pkg{dMod} provides also the \code{eqnlist} class which encodes the ODE by a list of (1) state names, (2) rate expressions, (3) compartment volumes and (4) the stoichiometric matrix. Each reaction flux, c.f.~eq.~\eqref{eq:Sv}, occurs only once and the gain and loss is represented by the coefficients in the stoichiometric matrix. \pkg{dMod} supports the read-in of a csv file with the stoichiometric matrix and rate expressions and provides a function \code{addReaction()} to construct an \code{eqnlist} object step-by-step or add further reactions to an already existing \code{eqnlist}. An \code{eqnlist} is converted to (differential) equations, i.e., an \code{eqnvec}, by \code{as.eqnvec()}. The conversion does all the bookkeeping of gain and loss terms and volume ratios due to compartment transitions.

@@

\pkg{dMod} seeks to stay close to the mathematical formulation, i.e., \code{x <- Xs(myodemodel)} will indeed return an \proglang{R} function, an object of class \code{prdfn}, which expects arguments \code{times} and \code{pars} and turns them into a model prediction. The letter \code{"s"} in \code{"Xs"} refers to sensitivities, i.e., the solution $x(t)$ is returned, the sensitivities $\frac{\partial x}{\partial p}(t)$ are returned, too.

Besides parameters, the prediction might also depend on forcings $u(t)$ and sudden events, e.g., setting states to 0 at a predefined point in time. Since forcings and events are by default fixed, they are defined together with the  prediction function, \code{x <- Xs(odemodel, forcings = myforcings, events = myevents)}. The definitions of both, forcings and events, are compatible with the way they are defined in \pkg{deSolve}.

@@

It is one of the fundamental concepts of \pkg{dMod} to allow concatenation of functions via the \code{"*"} operator. The mathematical formulation $y = g\circ x$, i.e., $y(t, p) = (g\circ x)(t, p):= g(x(t, p), p, t)$ becomes \code{y = g*x} in \proglang{R}. To obtain derivatives $\frac{\partial y}{\partial p}$, the chain rule is applied: $\frac{\partial y}{\partial p} = \frac{\partial g}{\partial x}\frac{\partial x}{\partial p} + \frac{\partial g}{\partial p}$. This means that \code{Y()} needs to be informed which symbols are states and parameters to generate the corresponding expressions $\frac{\partial g}{\partial x}$ and $\frac{\partial g}{\partial p}$. The observation function \code{g} takes care of computing theses derivatives from the symbolic expressions and doing the matrix multiplication with the sensitivities $\frac{\partial x}{\partial p}$ from the prediction function. Evaluation of symbolic expressions can become inefficient in \proglang{R}. Therefore, the observation function is usually translated into a \proglang{C} code and compiled.

When observation- and prediction functions are concatenated, the result is a prediction function, e.g., \code{y = g*x} is the \proglang{R} function computing values $y(t)$ from the arguments \code{times} and \code{pars} via evaluation of the ODE and subsequent evaluation of the observation function. Two observation functions can be concatenated, too, again yielding an observation function.

@@

Similar to prediction- and observation functions, parameter functions not only return parameter values but the Jacobian of the transformation, too. Exploiting the chain rule, the derivatives are propagated, allowing to define \code{y = g*x*p} which is a function returning $y(t, \theta)$ and $\frac{\partial y}{\partial \theta}(t)$, i.e., \code{g*x*p} is a prediction function.

@@

The aim of \pkg{dMod} is to allow for ``simultaneous'' predictions under several conditions and compare these predictions to the corresponding experimental data sets to estimate model parameters.
Different experimental conditions are typically expressed by the fact that some parameters are different between conditions whereas others are common to all conditions. This situation occurs, e.g., if perturbation experiments are performed, affecting only few parts of a system.
%The full information contained in several perturbation experiments is only used if the multi-conditional model prediction is set up in a way that all model parameters not being changed from condition to condition correspond to exactly one parameter. On the other hand, parameters which are systematically perturbed correspond to different parameters.
Mathematically speaking, we want to construct a parameter transformation
\begin{equation}
	\begin{aligned}
		\Phi:\quad \mathbb R^p & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^q \\
		\theta & \longmapsto  \left(\Phi_1(\theta), \dots, \Phi_n(\theta)\right)
	\end{aligned}
	\label{eq:partrafo}
\end{equation}
where $i = 1, \dots, n$ corresponds to the different conditions. If all parameters are shared throughout all conditions, then $p = q$. If, however, all parameters are distinct, then $p = n\cdot q$. Perturbation experiments correspond to a situation where $p\gtrsim q$.

@@

Prediction- and observation functions are generalized to multiple conditions by the \code{"+"} operator, too. Mathematically speaking, they become
\begin{align*}
	\begin{aligned}
		x: &  & \mathbb R \times \bigoplus_{i = 1}^n \mathbb R^q & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^m \\
		 & & (t, p_1, \dots, p_n) & \longmapsto (x_1(t, p_1), \dots, x_n(t, p_n)) \\
		g: &  & \bigoplus_{i = 1}^n \mathbb R^m \times \bigoplus_{i = 1}^n \mathbb R^q\times \mathbb R& \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^s \\
		& & (x_1(t), \dots, x_n(t), p_1, \dots, p_n, t) & \longmapsto (g_1(x_1(t), p_1, t), \dots, g_n(x_n(t, p_n), p_n, t)).
	\end{aligned}
	\label{}
\end{align*}
In words, if prediction- or observation functions are defined for different conditions then they expect condition-specific inputs which are evaluated by the matching functions. Examples for condition-specific prediction functions typically involve different forcings or events. Observation functions can, e.g., differ between different measurement techniques. In all these cases, the different prediction functions $x_1, \dots, x_n$ or observation functions $g_1, \dots, g_n$ are defined, referencing the condition, and combined by the \code{"+"} operator analogously to \code{p}.
% Aufruf der Funktionen mit subset von conditions

@@

\end{CodeInput}
\end{CodeChunk}

The events for \code{TCA_buffer} and \code{switch} are generated as part of the model and, therefore, need to be provided to the \code{odemodel()} command together with the model.
